Paper 3 - Causal Roadmap

### 2023-03-01 docs/causal_roadmap.qmd
Started working on DAGs: all nodes, DAG for chemical to intelligence
Cleaned the causal roadmap.

---
title: "Causal Roadmap"
author: Lorenzo Fabbri
output:
  bookdown::pdf_document2:
    number_sections: false
  html_document: default
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
```

```{r}
#| echo: false
# Type of adjustment set from DAG
type.mas <- "minimal"
# Type of effect to be estimated from DAG
type.effect <- "direct"
```

\textcolor{red}{Most of the following was **copy-pasted** from published papers. Some of the Sections must be filled depending on the chosen causal framework.}

@@

For time index $t$, let $W_t$ denote the set of potential confounders, $A_t$ the observed exposures, 
and $Y_t$ the clinical outcome of interest. For each subject $i$ and each time index $t$, 
we assumed its observed data $O_{ti} = (W_{ti}, A_{ti}, Y_{ti})$ were generated by sampling from a distribution 
$\mathbb{P}_{0,t}$ compatible with the causal model specified in \@ref(eq:npsem).

@@

\begin{equation}
\begin{aligned}
  W_t &= f_{W{_t}} (U_{W_{t}}) \\
  A_t &= f_{A_{t}} (W_t, U_{A_{t}}) \\
  Y_t &= f_{Y_{t}} (W_t, A_t, U_{Y_{t}}), 
\end{aligned}
(\#eq:npsem)
\end{equation}

where $(f_{W{_t}}, f_{A_{t}}, f_{Y_{t}})$ were the nonparametric structural equations, 
and $(U_{W_{t}}, U_{A_{t}}, U_{Y_{t}})$ were the unmeasured factors contributing the confounders, exposures, and outcome, respectively. 
The corresponding causal graph (without indexes for ease of notation) is given in Figure \@ref(fig:causal-graph).

```{r}
#| label: causal-graph
#| echo: false
#| fig.cap: Causal graph corresponding to structural equations.
#| fig.height: 3.0
#| fig.width: 3.0
causal.graph <- ggdag::dagify(A ~ W + U_A, Y ~ W + A + U_Y, W ~ U_W, 
                              exposure = "A", outcome = "Y") %>%
  ggdag::tidy_dagitty()
ggdag::ggdag(causal.graph) + ggdag::theme_dag()
```

@@

As **secondary analyses**, we estimated...

@@

```{r}
#| echo: false
#| fig.cap: Simplified DAG for research question 1.
sdag1 <- ggdag::dagify(X_ti ~ W_t, 
                       Y ~ W_t + X_ti, 
                       exposure = "X_ti", outcome = "Y") %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag(text_size = 2, node_size = 10) + ggdag::theme_dag() +
  ggplot2::ggtitle("RQ1")
```

```{r}
#| echo: false
ggpubr::ggarrange(sdag1, ..., 
                  ncol = x, nrow = y)
```

@@

```{r}
#| echo: false
source("../DAGs/dag_v1.R")
source("../code/dags.R")

dags <- dag.v1()
```

@@

```{r}
#| echo: false
#| fig.cap: DAG for Research question 1.
#| fig.height: 10.0
#| fig.width: 18.0
dag.rq1 <- dags$rq1 %>%
  dagitty::dagitty()
dag.rq1 %>% ggdag::tidy_dagitty() %>%
  ggdag::ggdag_status(text_size = 10, text_col = "black") +
  ggdag::theme_dag() + ggdag::scale_adjusted() +
  ggplot2::theme(legend.text = ggplot2::element_text(size = 20))
```

The `r I(type.mas)` sufficient adjustment sets for estimating the `r I(type.effect)` effect of X on Y are:

```{r}
#| echo: false
dagitty::adjustmentSets(x = dag.rq1, 
                        type = type.mas, effect = type.effect)
```

>>>>

---
title: "Paper 3 - Causal Roadmap"
keywords:
  - Causal Inference
  - Epidemiology
  - Triangulation
  - Modified Treatment Policy
  - Mendelian Randomization
  - Metabolomics
  - Endocrine Disrupting Chemicals
  - Neurodevelopment
  - Neurobehavior

author:
  - name: "Lorenzo Fabbri"
    number: 1
    email: lorenzo.fabbri@isglobal.org
    orcid: 0000-0003-3031-322X
    attributes:
      corresponding: true
    affiliations:
      - ref: isg
      - ref: upf
affiliations:
  - id: isg
    name: Barcelona Institute for Global Health (ISGlobal)
    city: Barcelona
    state: Spain
  - id: upf
    name: Pompeu Fabra University
    city: Barcelona
    state: Spain

format:
  pdf:
    documentclass: report
    pdf-engine: pdflatex
    toc: true
    margin-left: 15mm
    margin-right: 15mm
    margin-bottom: 15mm
    margin-top: 15mm

bibliography: ../output/paper/bibliography.bib
link-citations: true
linkcolor: blue

header-includes:
  - \usepackage{rotating}
  - \usepackage{float}

execute:
  echo: false

editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
```

```{r}
# Type of adjustment set from DAG
type_mas <- "minimal"
# Type of effect to be estimated from DAG
type_effect <- "direct"
```

\textcolor{red}{Most of the following was **copy-pasted** from published papers.}

@@

For time index $t$, let $W_t$ denote the set of potential confounders, $A_t$ the observed exposures, 
and $Y_t$ the clinical outcome of interest. For each subject $i$ and each time index $t$, 
we assumed its observed data $O_{ti} = (W_{ti}, A_{ti}, Y_{ti})$ were generated by sampling from a distribution 
$\mathbb{P}_{0,t}$ compatible with the causal model specified in @eq-npsem.

@@

$$
\begin{aligned}
  W_t &= f_{W{_t}} (U_{W_{t}}) \\
  A_t &= f_{A_{t}} (W_t, U_{A_{t}}) \\
  Y_t &= f_{Y_{t}} (W_t, A_t, U_{Y_{t}}), 
\end{aligned}
$$ {#eq-npsem}

where $(f_{W{_t}}, f_{A_{t}}, f_{Y_{t}})$ were the nonparametric structural equations, 
and $(U_{W_{t}}, U_{A_{t}}, U_{Y_{t}})$ were the unmeasured factors contributing the confounders, exposures, and outcome, respectively. 
The corresponding causal graph (without indexes for ease of notation) is given in @fig-causal-graph.

```{r}
#| label: fig-causal-graph
#| fig.cap: Causal graph corresponding to structural equations.
#| fig.height: 3.0
#| fig.width: 3.0
causal_graph <- ggdag::dagify(A ~ W + U_A, Y ~ W + A + U_Y, W ~ U_W, 
                              exposure = "A", outcome = "Y") %>%
  ggdag::tidy_dagitty()
ggdag::ggdag(causal_graph) + ggdag::theme_dag()
```

@@

@@

```{r}
#| fig.cap: Simplified DAG for research question 1.
sdag1 <- ggdag::dagify(X_ti ~ W_t, 
                       Y ~ W_t + X_ti, 
                       exposure = "X_ti", outcome = "Y") %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag(text_size = 2, node_size = 10) + ggdag::theme_dag() +
  ggplot2::ggtitle("RQ1")
```

## Research question 2:...

```{r}
#| fig.cap: Simplified DAG for research question 2.
sdag2 <- ggdag::dagify(X_ti ~ W_t, 
                       Y ~ W_t + X_ti, 
                       exposure = "X_ti", outcome = "Y") %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag(text_size = 2, node_size = 10) + ggdag::theme_dag() +
  ggplot2::ggtitle("RQ2")
```

## Research question 3:...

```{r}
#| fig.cap: Simplified DAG for research question 3.
sdag3 <- ggdag::dagify(X_ti ~ W_t, 
                       Y ~ W_t + X_ti, 
                       exposure = "X_ti", outcome = "Y") %>%
  ggdag::tidy_dagitty() %>%
  ggdag::ggdag(text_size = 2, node_size = 10) + ggdag::theme_dag() +
  ggplot2::ggtitle("RQ3")
```

```{r}
ggpubr::ggarrange(sdag1, sdag2, sdag3, 
                  ncol = 3, nrow = 1)
```

@@

```{r}
#| message: false
source("../DAGs/dag_v1.R")
source("../code/dags.R")

dags <- dag_v1()
```

@@

```{r}
#| fig.cap: DAG for Research question 1.
#| fig.height: 10.0
#| fig.width: 18.0
dag_rq1 <- dags$chem_to_out %>%
  dagitty::dagitty()
dag_rq1 %>% ggdag::tidy_dagitty() %>%
  ggdag::ggdag_status(text_size = 10, text_col = "black") +
  ggdag::theme_dag() + ggdag::scale_adjusted() +
  ggplot2::theme(legend.text = ggplot2::element_text(size = 20))
```

The `r I(type_mas)` sufficient adjustment sets for estimating the `r I(type_effect)` effect of X on Y are:

```{r}
dagitty::adjustmentSets(x = dag_rq1, 
                        type = type_mas, effect = type_effect)
```
